---
title: "Estimating the conditional variance by local linear regression"
author: "Caballero Vergés Biel, Menzenbach Svenja and Reyes Illescas Kleber Enrique"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r loading libraries, include=FALSE}
source("locpolreg.R")
library(sm) # for function "sm.regression"
library(KernSmooth) # for function "dpill"
```

```{r loading data, include=FALSE}
data(aircraft)

attach(aircraft)
lg_power <- log(Power)
lg_span <- log(Span)
lg_length <- log(Length)
lg_weight <- log(Weight)
lg_speed <- log(Speed)
lg_range <- log(Range)

```


```{r k-fold-cv, include=FALSE}
k.fold.cv <- function(x,y,k=10,h=range(x)/10,p=1,type.kernel="normal"){
  n <- length(x)
  Ik <- floor((0:(n-1))/(n/k))+1
  ssr <- 0
  for (i in (1:k)){
    y.i <- y[Ik==i]
    aux <- locpolreg(x[Ik!=i],y[Ik!=i],h=h,p=p,tg=x[Ik==i],
                     type.kernel=type.kernel, doing.plot=FALSE)
    ssr <- ssr + sum((y.i-aux$mtgr)^2)
  }
  k.cv <- ssr/n
  return(k.cv)
}

h.k.fold.cv <- function(x,y,h.v = exp(seq(log(diff(range(x))/20),
                                          log(diff(range(x))/4),l=12)), 
                        k=10,p=1,type.kernel="normal"){
  n <- length(x)
  perm <- sample(1:n)
  xperm <- x[perm]
  yperm <- y[perm]
  
  k.cv <- h.v*0
  for (i in (1:length(h.v))){
    h <- h.v[i]
    k.cv[i] <- k.fold.cv(x=xperm,y=yperm,k=k,h=h,p=p,
                         type.kernel=type.kernel)
  }
  return(list(k=k,h.v=h.v,k.cv=k.cv))
}
```


```{r fit llr, include=FALSE}
fit.loc_pol_reg <- function(x, y){
  out.cv <- h.k.fold.cv(x=x, y=y, k=n) # this function uses locpolreg
  h_opt <- out.cv$h.v[which.min(out.cv$k.cv)]
  return <- (list(lpr=locpolreg(x=x, y=y, h=h_opt, doing.plot=TRUE), 
                  h_opt=h_opt))
}
```

```{r variables}
x <- Yr
y <- lg_weight
n <- length(x)
```

# Estimating the conditional variance (Using loc.pol.reg)

## 1.Fit a nonparametric regression to data $(x_i, y_i)$ and save the estimated values $\hat{m}(x_i)$.
```{r question 1}
fit.lpr <- fit.loc_pol_reg(x, y)
fit_y <- fit.lpr$lpr
mtgr <- fit_y$mtgr # \hat_{m}(x_i)

```

## 2.Transform the estimated residuals $\hat{\epsilon} = y_i - \hat{m}(x_i)$
\[
z_i = \log \epsilon_i^2 = log((y_i - \hat{x_i}))^2)
\] 
```{r question 2}
hat_e <- y - mtgr
z <- log((hat_e)^2)
```

## 3. Fit a nonparametric regression to data $(x_i, z_i)$ and call the estimated function $\hat{q}(x)$. Observe that $\hat{q}(x)$ is an estimate of log $\sigma^2(x)$.
```{r question 3}
fit_z <- fit.loc_pol_reg(x, z)$lpr
qtgr <- fit_z$mtgr
```

## 4. Estimate $\sigma^2(x)$ by 
\[
\hat{\sigma}^2(x) = e^{\hat{q}(x)}
\]
```{r question 4}
sig.sqr <- exp(qtgr)
```

## plots
```{r plot 1}
plot(x, (hat_e)^2)
lines(x, sig.sqr, col=4, lwd=3)
```

```{r plot 2}
y.min <- min(mtgr - 1.96 * sqrt(sig.sqr)) - 0.1
y.max <- max(mtgr + 1.96 * sqrt(sig.sqr)) + 0.1
plot(x,mtgr, type = 'l', ylim=c(y.min, y.max))
lines(x,mtgr + 1.96 * sqrt(sig.sqr), col="blue")
lines(x,mtgr - 1.96 * sqrt(sig.sqr), col="blue")
```
# Estimating the conditional variance (Using sm.regression)

## 1.Fit a nonparametric regression to data $(x_i, y_i)$ and save the estimated values $\hat{m}(x_i)$.
```{r question 1_2}
h.dpi <- dpill(x=x, y=y, range.x=range(x))
sm_regression <- sm.regression(x=x, y=y, eval.points=x, h=h.dpi, pch=1, cex=1, col=2, lwd=2)
mtgr_sm <- sm_regression$estimate
```

## 2.Transform the estimated residuals $\hat{\epsilon} = y_i - \hat{m}(x_i)$
\[
z_i = \log \epsilon_i^2 = log((y_i - \hat{x_i}))^2)
\] 
```{r question 2_2}
hat_e_sm <- y - mtgr_sm
z_sm <- log((hat_e_sm)^2)
```

## 3. Fit a nonparametric regression to data $(x_i, z_i)$ and call the estimated function $\hat{q}(x)$. Observe that $\hat{q}(x)$ is an estimate of log $\sigma^2(x)$.
```{r question 3_2}
sm_regression_z <- sm.regression(x=x, y=z, pch=1, cex=1, col=2, lwd=2,
                               eval.points=x, 
                               h=dpill(x=x, y=z, range.x=range(x)))
qtgr_sm <- sm_regression_z$estimate
```

## 4. Estimate $\sigma^2(x)$ by 
\[
\hat{\sigma}^2(x) = e^{\hat{q}(x)}
\]
```{r question 4_2}
sig.sqr_sm <- exp(qtgr_sm)
```

## plots
```{r plot 1_2}
plot(x, (hat_e_sm)^2)
lines(x, sig.sqr_sm, col=4, lwd=3)
```

```{r plot 2_2}
y.min_sm <- min(mtgr_sm - 1.96 * sqrt(sig.sqr_sm)) - 0.1
y.max_sm <- max(mtgr_sm + 1.96 * sqrt(sig.sqr_sm)) + 0.1
plot(x,mtgr_sm, type = 'l', ylim=c(y.min_sm, y.max_sm))
lines(x,mtgr_sm + 1.96 * sqrt(sig.sqr_sm), col="blue")
lines(x,mtgr_sm - 1.96 * sqrt(sig.sqr_sm), col="blue")
```