---
title: "Estimating the conditional variance by local linear regression"
author: "Caballero Vergés Biel, Menzenbach Svenja and Reyes Illescas Kleber Enrique"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r loading libraries, include=FALSE}
library(sm)
```

```{r loading data, include=FALSE}
data(aircraft)
help(aircraft)
attach(aircraft)
lgPower <- log(Power)
lgSpan <- log(Span)
lgLength <- log(Length)
lgWeight <- log(Weight)
lgSpeed <- log(Speed)
lgRange <- log(Range)
```

# Estimating the conditional variance

## 1.Fit a nonparametric regression to data $(x_i, y_i)$ and save the estimated values $\hat{m}(x_i)$.
```{r question 1, echo=FALSE}
```

## 2.Transform the estimated residuals $\hat{\epsilon} = y_i - \hat{m}(x_i)$
\[
z_i = \log \epsilon_i^2 = log((y_i - \hat{x_i}))^2)
\] 
```{r question 2, echo=FALSE}
```

## 3. Fit a nonparametric regression to data $(x_i, z_i)$ and call the estimated function $\hat{q}(x)$. Observe that $\hat{q}(x)$ is an estimate of log $\sigma^2(x)$.
```{r question 3, echo=FALSE}
```

## 4. Estimate $\sigma^2(x)$ by 
\[
\hat{\sigma}^2(x) = e^{\hat{q}(x)}
\]
```{r question 4, echo=FALSE}
```