---
title: "Local Poisson regression"
author: "Biel Caballero Vergés, Svenja Menzenbach and Kleber Enrique Reyes Illescas"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r loading libraries, include=FALSE}
library(sm) # for function "sm.poisson"
```

# 1. Bandwidth choice for the local Poisson regression

In this section, we have modify the *h.cv.sm.binomial* (now *h.cv.sm.poisson*) and the *loglik.CV* functions in order to be able to use it for the local Poisson regression. We have simplified the code to have less parameters. 
```{r}
h.cv.sm.poisson <- function(x, y, rg.h = NULL, l.h = 10) {
  cv.h <- numeric(l.h)
  
  if (is.null(rg.h)) {
    hh <- c(h.select(x, y, method = "cv"), h.select(x, y, method = "aicc"))
    rg.h <- range(hh) * c(1/1.1, 1.5)
  }
  
  gr.h <- exp(seq(log(rg.h[1]), log(rg.h[2]), length.out = l.h))
  
  for (i in 1:length(gr.h)) {
    cv.h[i] <- loglik.CV(x, y, gr.h[i])
  }
  
  return(list(h = gr.h, cv.h = cv.h, h.cv = gr.h[which.max(cv.h)]))
}
```

```{r}
loglik.CV <- function(x, y, h) {
    n <- length(x)
    
    lambda <- sapply(1:n, 
                 function(i, x, y, h) {
        sm.poisson(x = x[-i], y = y[-i], h = h, 
                   eval.points = x[i], display = "none")$estimate
    }, x, y, h)
    
    return(sum(-lambda + y*log(lambda) -log(factorial(y)) )/n )
}
```

\newpage

# 2. Local Poisson regression for Country Development Data

With our new function, we can acquire an optimal bandwidth for fitting our model.
```{r}

countries <- read.csv2(file="HDI.2017.subset.csv",row.names = 1)

life.expec <- countries$Life.expec
le.fm.r <- round(countries$le.fm)

cv_result <- h.cv.sm.poisson(life.expec, le.fm.r, rg.h=c(0.5,10), l.h = 30)

plot(cv_result$h, cv_result$cv.h)
selected.bandwidth <- cv_result$h.cv
abline(v = selected.bandwidth, col="8", lty=2)
mid.point <- (max(cv_result$cv.h)+min(cv_result$cv.h))/2
text(selected.bandwidth, mid.point, 
     round(selected.bandwidth*100)/100,cex=0.65, pos=4,col=1) 
lines(cv_result$h, cv_result$cv.h)

model <- sm.poisson(x = life.expec, y = le.fm.r, h = selected.bandwidth, col=1)

```
The leave-one-out cross-validation determines 2.61 as optimal bandwidth. The fit of the local Poisson regression looks smooth and reasonable. It follows the course of high density of points while generalizing to all data. We do not observe overfitting.
