---
title: "Density Estimation"
subtitle: "Bandwidht choice by leave-one-out maximum likelihood"
author: "Biel Caballero, Menzenbach Svenja and Reyes Illescas Kleber Enrique"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Histogram

1. At the slides we have seen the following relationship \[\hat{f}_{h,(-i)}(x_i) =  \frac{n}{n-1} \left( \hat{f}_{h}(x_i) -\frac{K(0)}{nh} \right)\] between the leave-one-out kernel density estimator $\hat{f}_{h,(-i)}(x)$ and the kernel density estimator using all the observations $\hat{f}_{h}(x)$, when both are evaluated at $x_i$, one of the observed data. Find a similar relationship between the histogram estimator of the density function $\hat{f}_{hist}(x)$ and its leave-one-out version, $\hat{f}_{hist,(-i)}(x)$, when both are evaluated at $x_i$.

2. Read the CD rate data set and call x the first column. Then define A, Z and nbr and plot the histogram of x
```{r CDrate}
cdrate.df <-read.table("./cdrate.dat.txt")
# head(cdrate.df)
x <- cdrate.df[,1]
# sort(CDrate)
# # Stem-and-Leaf plot
# stem(CDrate)
```


```{r}
A <- min(x)-.05*diff(range(x)) 
Z <- max(x)+.05*diff(range(x))
nbr <- 7

hx <- hist(x,breaks=seq(A,Z,length=nbr+1),freq=F)
hx_f <- stepfun(hx$breaks,c(0,hx$density,0))
points(x, hx_f(x))
```


```{r}


```



3. Use the formula you have found before relating $\hat{f}_{hist}(x_i)$ and $\hat{f}_{hist,(-i)}(x_i)$ to compute $\hat{f}_{hist,(-i)}(x),i=1,...,n$, . Then add the points $(x_i, \hat{f}_{hist,(-i)}(x_i)),i=1,...,n$,  , to the previous plot
```{r}
hx_f2<-length(x)/(length(x)-1)*hx_f(x)-1/((length(x)-1)*(hx$breaks[2]-hx$breaks[1]))

A <- min(x)-.05*diff(range(x)) 
Z <- max(x)+.05*diff(range(x))
nbr <- 7
hx <- hist(x,breaks=seq(A,Z,length=nbr+1),freq=F)
hx_f <- stepfun(hx$breaks,c(0,hx$density,0))
points(x, hx_f(x))
points(x, hx_f2, col="red")
```

4. Compute the leave-one-out log-likelihood function corresponding to the previous histogram, at which *nbr=7* has been used
```{r}
looCV_log_lik_7 <- sum(log(hx_f2))
looCV_log_lik_7
```


5. **Choosing** *nbr* **by leave-one-out Cross Validation (looCV)**. Consider now the set *seq(1,15)* as possible values for *nbr*, the number of intervals of the histogram. For each of them compute the leave-one-out log-likelihood function (*looCv_log_lik*) fir the corresponding histogram. Then plot the values of *looCv_log_lik* against the values of *nbr* and select the optimal value of *nbr* as that at which *looCv_log_lik* takes its maximum. Finally, plot the histogram of *x* using the optimal value of *nbr*
```{r}
#sum of the product of the hx_f2 vector plot histograms for different number of breaks nbr

log_liks = list()

for (nbr in c(1:15)){
  #A <- min(hx_f2)-.05*diff(range(hx_f2)) 
  #Z <- max(hx_f2)+.05*diff(range(hx_f2))
  hx_i <- hist(x,breaks=seq(A,Z,length=nbr+1),freq=F, plot = FALSE)
  hx_f_i <- stepfun(hx_i$breaks,c(0,hx_i$density,0))
  n <- length(x)
  b <- hx_i$breaks[2]-hx_i$breaks[1]
  hx_f2_i <- n/(n-1)*hx_f_i(x) - 1/((n-1)*b)
  print(min(hx_f2_i))
  
  hx_i <- hist(x,breaks=seq(A,Z,length=nbr+1),freq=F)
  points(x, hx_f_i(x))
  points(x, hx_f2_i, col='red')

  looCV_log_lik <- sum(log(hx_f2_i))
  log_liks <- append(log_liks, looCV_log_lik)
}
plot(seq(1, 15), log_liks)
nbr_opt <- which.max(log_liks)
hist(x,breaks=seq(A,Z,length=nbr_opt+1),freq=F)

# TODO: - avoid production of NaNs, 
#       - is it a problem that 'freq' is not used for some cases?

```

6. **Chossing** *b* **by looCV**. Let *b* be the common width of the bins of a histogram. Consider the set $seq((Z-A)/15, (Z-A)/1, length=30)$ as possible values for *b*. Select the value of *b* maximizing the leave-one-out log-likelihood function, and plot the corresponding histogram
```{r}
b_set <- seq((Z-A)/15, (Z-A), length=30)

for(b in b_set){
  hx <- hist(x, breaks=seq(A, Z+b, by=b), plot=F)
  plot(hx, freq = FALSE)
}

```
7. Recycle the functions *graph.mixt* and *sim.mixt* defined at *density_estimation.Rmd* to generate n = 100 data from \[f(x)=(3/4)N(x;m=0,s=1)+(1/4)N(x;m=3/2,s=1/3)\] Let *b* be the bin width of a histogram estimator of f(x) using the generated data. Select the value of *b* maximizing the leave-one-out log-likelihood function, and plot the corresponding histogram. Compare with the results obtained using the Scott’s formula: \[b_{Scott} = 3.49St.Dev(X)_n^{-1/3}\].

```{r}
# TODO

```



# Kernel density estimator

8.\[\hat{f}_{h,(-i)}(x_i) =  \frac{n}{n-1} \left( \hat{f}_{h}(x_i) -\frac{K(0)}{nh} \right)\]



























