---
title: "Density Estimation"
subtitle: "Bandwidht choice by leave-one-out maximum likelihood"
author: "Biel Caballero, Menzenbach Svenja and Reyes Illescas Kleber Enrique"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Histogram

1. At the slides we have seen the following relationship \[\hat{f}_{h,(-i)}(x_i) =  \frac{n}{n-1} \left( \hat{f}_{h}(x_i) -\frac{K(0)}{nh} \right)\] between the leave-one-out kernel density estimator $\hat{f}_{h,(-i)}(x)$ and the kernel density estimator using all the observations $\hat{f}_{h}(x)$, when both are evaluated at $x_i$, one of the observed data. Find a similar relationship between the histogram estimator of the density function $\hat{f}_{hist}(x)$ and its leave-one-out version, $\hat{f}_{hist,(-i)}(x)$, when both are evaluated at $x_i$.

2. Read the CD rate data set and call x the first column. Then define A, Z and nbr and plot the histogram of x
```{r CDrate}
cdrate.df <-read.table("./cdrate.dat.txt")
# head(cdrate.df)
x <- cdrate.df[,1]
# sort(CDrate)
# # Stem-and-Leaf plot
# stem(CDrate)
```


```{r}
A <- min(x)-.05*diff(range(x)) 
Z <- max(x)+.05*diff(range(x))
nbr <- 7

hx <- hist(x,breaks=seq(A,Z,length=nbr+1),freq=F)
hx_f <- stepfun(hx$breaks,c(0,hx$density,0))
points(x, hx_f(x))
```


```{r}


```



3. Use the formula you have found before relating $\hat{f}_{hist}(x_i)$ and $\hat{f}_{hist,(-i)}(x_i)$ to compute $\hat{f}_{hist,(-i)}(x),i=1,...,n$, . Then add the points $(x_i, \hat{f}_{hist,(-i)}(x_i)),i=1,...,n$,  , to the previous plot
```{r}
hx_f2<-length(x)/(length(x)-1)*hx_f(x)-1/((length(x)-1)*(hx$breaks[2]-hx$breaks[1]))

A <- min(x)-.05*diff(range(x)) 
Z <- max(x)+.05*diff(range(x))
nbr <- 7
hx <- hist(x,breaks=seq(A,Z,length=nbr+1),freq=F)
hx_f <- stepfun(hx$breaks,c(0,hx$density,0))
points(x, hx_f(x))
points(x, hx_f2, col="red")
```

4. Compute the leave-one-out log-likelihood function corresponding to the previous histogram, at which *nbr=7* has been used
```{r}
looCV_log_lik_7 <- sum(log(hx_f2))
looCV_log_lik_7
```


5. **Choosing** *nbr* **by leave-one-out Cross Validation (looCV)**. Consider now the set *seq(1,15)* as possible values for *nbr*, the number of intervals of the histogram. For each of them compute the leave-one-out log-likelihood function (*looCv_log_lik*) fir the corresponding histogram. Then plot the values of *looCv_log_lik* against the values of *nbr* and select the optimal value of *nbr* as that at which *looCv_log_lik* takes its maximum. Finally, plot the histogram of *x* using the optimal value of *nbr*
```{r}
#sum of the product of the hx_f2 vector plot histograms for different number of breaks nbr

log_liks = list()

for (nbr in c(1:15)){
  #A <- min(hx_f2)-.05*diff(range(hx_f2)) 
  #Z <- max(hx_f2)+.05*diff(range(hx_f2))
  hx_i <- hist(x,breaks=seq(A,Z,length=nbr+1),freq=F, plot = FALSE)
  hx_f_i <- stepfun(hx_i$breaks,c(0,hx_i$density,0))
  n <- length(x)
  b <- hx_i$breaks[2]-hx_i$breaks[1]
  hx_f2_i <- n/(n-1)*hx_f_i(x) - 1/((n-1)*b)
  print(min(hx_f2_i))
  
  hx_i <- hist(x,breaks=seq(A,Z,length=nbr+1),freq=F)
  points(x, hx_f_i(x))
  points(x, hx_f2_i, col='red')

  looCV_log_lik <- sum(log(hx_f2_i))
  log_liks <- append(log_liks, looCV_log_lik)
}
plot(seq(1, 15), log_liks)
nbr_opt <- which.max(log_liks)
hist(x,breaks=seq(A,Z,length=nbr_opt+1),freq=F)

# TODO: - avoid production of NaNs, 
#       - is it a problem that 'freq' is not used for some cases?

```

6. **Chossing** *b* **by looCV**. Let *b* be the common width of the bins of a histogram. Consider the set $seq((Z-A)/15, (Z-A)/1, length=30)$ as possible values for *b*. Select the value of *b* maximizing the leave-one-out log-likelihood function, and plot the corresponding histogram
```{r}
b_set <- seq((Z-A)/15, (Z-A), length=30)

b_max_log_lik <- function(b_set, x){
  
  log_liks_b = list()
  
  for(b in b_set){
    hx <- hist(x, breaks=seq(A, Z+b, by=b), plot=F)
    
    # compute histogram estimator
    hx_f <- stepfun(hx$breaks,c(0,hx$density,0))
    n <- length(x)
    hx_f2 <- n/(n-1)*hx_f_i(x) - 1/((n-1)*b)
    
    looCV_log_lik_b <- sum(log(hx_f2))
    log_liks_b <- append(log_liks, looCV_log_lik_b)
    
  }
  
  b_opt <- which.max(log_liks_b)
  b <- b_set[b_opt]
  return(b)
y}

b <- b_max_log_lik(b_set, x)

hx <- hist(x, breaks=seq(A, Z+b, by=b), plot=F)
plot(hx, freq = FALSE)

# TODO: avoid NaNs
```
7. Recycle the functions *graph.mixt* and *sim.mixt* defined at *density_estimation.Rmd* to generate n = 100 data from \[f(x)=(3/4)N(x;m=0,s=1)+(1/4)N(x;m=3/2,s=1/3)\] Let *b* be the bin width of a histogram estimator of f(x) using the generated data. Select the value of *b* maximizing the leave-one-out log-likelihood function, and plot the corresponding histogram. Compare with the results obtained using the Scott’s formula: \[b_{Scott} = 3.49St.Dev(X)_n^{-1/3}\].

```{r}
# graph.mixt
# Input:
#    k: number mixture components
#    mu: vector of length k with the mean values of the k normals
#    sigma: vector of length k with the st.dev. values of the k normals
#    alpha: vector of length k with the weights of each normal
#    graphic: logical value indicating if the mixture density must be plotted
#    ...: Other parameters passed to plot()
#
# Output:
#    L, U: extremes of the interval where the mixture density is plotted
#    x: points at which the mixture density is evaluated 
#    fx: value of the mixture density at x
#
graph.mixt<-
function(k=1, mu=seq(-2*(k-1),2*(k-1),length=k), sigma=seq(1,1,length=k), alpha=seq(1/k,1/k,length=k), graphic=TRUE,...)
{
   L<-min(mu-3*sigma)
   U<-max(mu+3*sigma)
         
   x<- seq(from=L,to=U,length=200)
   fx<- 0*x
   Salpha<-sum(alpha)
   for(i in 1:k){
    p<-alpha[i]/Salpha
#       fx <- fx + p*exp(-.5*((x-mu[i])/sigma[i])^2)/(sqrt(2*pi)*sigma[i])
    fx <- fx + p*dnorm(x,mu[i],sigma[i])
   }
   if (graphic){
      plot(x,fx,type="l",...)
   }
   return(list(L = L, U = U, x = x, fx = fx))
}

# sim.mixt
# Input:
#    n: number of simulated data
#    k: number mixture components
#    mu: vector of length k with the mean values of the k normals
#    sigma: vector of length k with the st.dev. values of the k normals
#    alpha: vector of length k with the weights of each normal
#    graphic: logical value indicating if the mixture density and the 
#              histogram of the simulated data must be plotted
#    ...: Other parameters passed to plot()
#
# Output:
#    x: simulated data
#
# Requires: 
#    graph.mixt
sim.mixt <- function(n=1,k=1, 
         mu=seq(-2*(k-1),2*(k-1),length=k), 
         sigma=seq(1,1,length=k), 
         alpha=seq(1/k,1/k,length=k), graphic=FALSE,...)
{
   csa<-cumsum(alpha)
   x<-runif(n)
      
   for (i in 1:n){
      comp<-sum(csa<=x[i])+1
      x[i]<-rnorm(1,mu[comp],sigma[comp])
   }
   if(graphic) {
      out<-graph.mixt(k, mu, sigma, alpha, gr=FALSE)
      hist(x,freq = FALSE,
           ylim=c(0,max(c(max(out$fx),max(hist(x,plot=FALSE)$density)))))
      lines(out$x,out$fx,lty=1,lwd=2)
   }   
   return(x)
}
```


```{r}
set.seed(123)
n <- 100
mu <- c(0,3/2)
sigma <- c(1,1/3)
alpha <- c(3/4,1/4)
x <- sim.mixt(n=n, k=2, mu=mu, sigma=sigma, alpha=alpha, gr=T)
points(x,0*x,pch="|")
```
```{r}
b_scott <- function(X){
  return(3.49 * sd(X) * length(X)**(-1/3))
}

A <- min(x)-.05*diff(range(x)) 
Z <- max(x)+.05*diff(range(x))

b <- b_max_log_lik(b_set, x)
hx <- hist(x, breaks=seq(A, Z+b, by=b), plot=F)
plot(hx, freq = FALSE)

b_s <- b_scott(x)
hx_s <- hist(x, breaks=seq(A, Z+b_s, by=b_s), plot=F)
plot(hx_s, freq = FALSE)


```



# Kernel density estimator

8.\[\hat{f}_{h,(-i)}(x_i) =  \frac{n}{n-1} \left( \hat{f}_{h}(x_i) -\frac{K(0)}{nh} \right)\]


```{r}
kx <- density(x)
kx_f <- approxfun(x=kx$x, y=kx$y, method = 'linear', rule=2)
```

























