---
title: "Non-linear dimensionality reduction"
subtitle: "Principal curves, local MDS, Isomap and t-SNE"
author: "Caballero Vergés Biel, Menzenbach Svenja and Reyes Illescas Kleber Enrique"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r loading libraries, include=FALSE}
library(plot3D)
library(stops)
library(Rtsne)
```


# PART B. Local MDS, ISOMAP and t-SNE

```{r PART B, include=FALSE}
# ploting 1 digit
plot.zip <- function(x,use.first=FALSE,...){
  x<-as.numeric(x)
  if (use.first){
    x.mat <- matrix(x,16,16)
  }else{
    x.mat <- matrix(x[-1],16,16)
  }
  image(1:16,1:16,x.mat[,16:1],
        col=gray(seq(1,0,l=12)),...)
  invisible(
    if (!use.first){
      title(x[1])
    }else{
    }
  )  
  #col=gray(seq(1,0,l=2)))
}
```


## 2. Local MDS for ZERO digits

```{r PART B reading data}
zip.train <- read.table("zip.train")
zip.train.0 <- zip.train[zip.train[1] == 0, -1]
```

```{r PART B-2 a}
row.names(zip.train.0) <- NULL
dist.zip.0 <- dist(zip.train.0)

k <- 5
tau <- .05
q<-2 # 2-dim config

conf0 <- stats::cmdscale(dist.zip.0, k=q)

if (!file.exists("lmds1.RData")) {
  lmds.S.res <- lmds(as.matrix(dist.zip.0), init=conf0, ndim=q, k=k, tau=tau, itmax = 1000)
} else {
  load("lmds1.RData")
}

conf.lmds.S.res <- lmds.S.res$conf
```

```{r PART B-2 a store, include=FALSE}
if (!file.exists("lmds1.RData")) {
  save(lmds.S.res, file="lmds1.RData")
}
```

```{r PART B-2 a2, out.width="90%", fig.align = "center"}
plot(conf.lmds.S.res, main=paste0("Local MDS, k=",k,", tau=",tau))
```
We've selected 9 points to encompass the data's variability. These points are marked in red, and each one has assigned a number in order to recognized each point with its representation using the *plot.zip* function.

```{r PART B-2 b, out.width="90%", fig.align = "center"}
#filt <- as.data.frame(conf.lmds.S.res)
#filt2 <- subset(filt, V1 > 40 & V1 < 55 & V2 < -7 & V2 > -14)
selected.points <- c(443, 575, 474, 291, 905, 731, 644, 242, 201)
aux <- conf.lmds.S.res[-selected.points,]
aux2 <- conf.lmds.S.res[selected.points,]

plot(aux, main=paste0("Local MDS, k=",k,", tau=",tau), col=8)
points(aux2, pch = 16, col=2)
text(aux2[,1],aux2[,2],1:9, pos=3, col=2, font = 2, cex = 1.5)
```

We can observe 2 patterns in the printed zeros below. The first one (the y-axis) is related to thickness of the stroke, positive values are thicker (the higher the value, the higher the stroke) and vice versa (lower values result in thinner strokes). On the other hand, the x-axis could be associated with the roundness of the number (negative values resemble an oval shape, while positive values resemble a circle shape). Furthermore, it appears that positive values in x-axis tend to result in unfinished strokes, like fast writing.


```{r PART B-2 b zero plots, fig.show="hold", out.width="32%"}
for (i in 1:length(selected.points)) {
  index <- selected.points[i]
  plot.zip(zip.train.0[index,], use.first=TRUE)
  title(i, cex.main = 2.5)
}

```


```{r PART B-2 c LCMC function, include=FALSE}
LCMC <- function(D1,D2,Kp){
  D1 <- as.matrix(D1)
  D2 <- as.matrix(D2)
  n <- dim(D1)[1]
  N.Kp.i <- numeric(n)
  for (i in 1:n){
    N1.i <- sort.int(D1[i,],index.return = TRUE)$ix[1:Kp]
    N2.i <- sort.int(D2[i,],index.return = TRUE)$ix[1:Kp]
    N.Kp.i[i] <- length(intersect(N1.i, N2.i))
  }
  N.Kp<-mean(N.Kp.i)
  M.Kp.adj <- N.Kp/Kp - Kp/(n-1)
  
  return(list(N.Kp.i=N.Kp.i, M.Kp.adj=M.Kp.adj))
}
```



```{r PART B-2 c}
#dist.zip.0 <- dist(zip.train.0)
q <- 2
Kp <- 10

K <- c(5,10,15)
Tau <- c(.1,.5,1)

if (!file.exists("lmds_k_tau.RData")) {
  LC <- matrix(0,nrow=length(K),ncol=length(Tau))
  lmds.k.tau <- array(vector("list",1),dim=dim(LC))
  
  for (i in 1:length(K)){
    for (j in 1:length(Tau)){
      lmds.k.tau[[i,j]] <- lmds(as.matrix(dist.zip.0), init=conf0, 
                                ndim=q, k=K[i], tau=Tau[j], itmax=1000)$conf
      D2.k.tau <- dist(lmds.k.tau[[i,j]])
      LC[i,j] <- LCMC(dist.zip.0,D2.k.tau,Kp)$M.Kp.adj
    }
    
  }
} else {
  load("lmds_k_tau.RData")
  load("LC.RData")
}


ij.max <- arrayInd(which.max(LC),.dim=dim(LC))
k.max <- K[ij.max[1]] 
tau.max <- Tau[ij.max[2]] 
lmds.max <- lmds.k.tau[[ij.max[1],ij.max[2]]]

print(paste0("k.max=",k.max,"; tau.max=",tau.max, "; LC.max=", LC[ij.max[1], ij.max[2]]))


```
```{r PART B-2 c store, include=FALSE}
if (!file.exists("LC.RData")) {
  save(LC, file="LC.RData")
}

if (!file.exists("lmds_k_tau.RData")) {
  save(lmds.k.tau, file="lmds_k_tau.RData")
}
```



```{r PART B-2 c plot, out.width="90%", fig.align = "center"}
selected.points <- c(344, 962, 221, 530, 399, 440)
aux <- lmds.max[-selected.points,]
aux2 <- lmds.max[selected.points,]

plot(aux, main=paste0("Local MDS, k=",k.max,", tau=",tau.max), col=8)
points(aux2, pch = 16, col=2)
text(aux2[,1],aux2[,2],1:6, pos=2, col=2, font = 2, cex = 1.5)
```

```{r PART B-2 c zero plot, fig.show="hold", out.width="30%"}
for (i in 1:length(selected.points)) {
  index <- selected.points[i]
  plot.zip(zip.train.0[index,], use.first=TRUE)
  title(i, cex.main=2.5)
}
```

## 3.
```{r PART B-3 a}


```



## 4.

In this last exercise we have tried t-sne 
```{r PART B-4 a}
#row.names(zip.train.0) <- NULL
#zip_matrix <- as.matrix(zip.train.0)

if (!file.exists("tsne_out.RData")) {
  tsne_out <- Rtsne(dist.zip.0, dims=2, pca=FALSE,perplexity=40,theta=0.0, max_iter=1000, num_threads=4)
} else {
  load("tsne_out.RData")
}
```

```{r PART B-4 a store, include=FALSE}
if (!file.exists("tsne_out.RData")) {
  save(tsne_out, file="tsne_out.RData")
}
```

```{r PART B-4 a plot, out.width="90%", fig.align = "center"}
plot(tsne_out$Y)
```
In this case we can notice than the y-axis tends to increase while x-axis increase, but the conclusions are similar to previous exercises. The y-axis is correlated with thickness and x-axis with roundness. We have to higlight the fact that in this case lower values are thickness but the overall interpretation is equal.
```{r PART B-4 b plot, out.width="90%", fig.align = "center"}
#filt <- as.data.frame(tsne_out$Y)
#filt2 <- subset(filt, V1 > -10 & V1 < 0 & V2 > -25 & V2 < -18)
selected.points <- c(443, 575, 474, 139, 489, 732, 198, 242, 699 )
tsne_out.no.selected <- tsne_out$Y[-selected.points,]
tsne_out.selected <- tsne_out$Y[selected.points,]

plot(tsne_out.no.selected, col=8)
points(tsne_out.selected, pch = 16, col=2)
text(tsne_out.selected[,1],tsne_out.selected[,2],1:9, pos=2, col=2, font = 2, cex = 1.5)
```


```{r PART B-4 b zero plot, fig.show="hold", out.width="30%"}
for (i in 1:length(selected.points)) {
  index <- selected.points[i]
  plot.zip(zip.train.0[index,], use.first=TRUE)
  title(i, cex.main=2.5)
}
```
We apply the LCMC with the perplexity parameter. 

```{r PART B-4 c}
Kp <- 10

perp <- c(10,15,20,25,30,35,40)

if (!file.exists("Rtsne_perp.RData")) {
  LC.Rtsne <- numeric(length(perp))
  Rtsne.perp <- vector("list",length(perp))
  
  for (i in 1:length(perp)){
    Rtsne.perp[[i]] <- Rtsne(dist.zip.0, dims=2, pca=FALSE, perplexity=perp[i],theta=0.0, max_iter=1000, num_threads=4)
    D2.perp <- dist(Rtsne.perp[[i]]$Y)
    LC.Rtsne[i] <- LCMC(dist.zip.0,D2.perp,Kp)$M.Kp.adj
    
  }
} else {
  load("LC_rtsne.RData")
  load("Rtsne_perp.RData")
}


i.max <- which.max(LC.Rtsne)
perplexity.max <- perp[i.max[1]] 
Rtsne.max <- Rtsne.perp[[i.max]]

plot(perp, LC.Rtsne, main=paste0("perplexity.max=",perplexity.max))
abline(v=perp[i.max],col=2)
```


```{r PART B-4 c store, include=FALSE}
if (!file.exists("LC_rtsne.RData")) {
  save(LC.Rtsne, file="LC_rtsne.RData")
}

if (!file.exists("Rtsne_perp.RData")) {
  save(Rtsne.perp, file="Rtsne_perp.RData")
}
```

In this part we used the same points as in B2c part. The conclusions are similars

```{r PART B-4 c plot, out.width="90%", fig.align = "center"}
selected.points <- c(344, 962, 221, 530, 399, 440)
aux <- Rtsne.max$Y[-selected.points,]
aux2 <- Rtsne.max$Y[selected.points,]

plot(aux, main=paste0("Rtsne, perplexity=",perplexity.max), col=8)
points(aux2, pch = 16, col=2)
text(aux2[,1],aux2[,2],1:6, pos=2, col=2, font = 2, cex = 1.5)
```

```{r PART B-4 c zero plot, fig.show="hold", out.width="30%"}
for (i in 1:length(selected.points)) {
  index <- selected.points[i]
  plot.zip(zip.train.0[index,], use.first=TRUE)
  title(i, cex.main=2.5)
}
```

## 5.
```{r PART B-5}


```




