---
title: "Non-linear dimensionality reduction"
subtitle: "Principal curves, local MDS, Isomap and t-SNE"
author: "Caballero Vergés Biel, Menzenbach Svenja and Reyes Illescas Kleber Enrique"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r loading libraries, include=FALSE}
library(plot3D)
library(stops)
library(Rtsne)
```


# PART B. Local MDS, ISOMAP and t-SNE

```{r PART B, include=FALSE}
# ploting 1 digit
plot.zip <- function(x,use.first=FALSE,...){
  x<-as.numeric(x)
  if (use.first){
    x.mat <- matrix(x,16,16)
  }else{
    x.mat <- matrix(x[-1],16,16)
  }
  image(1:16,1:16,x.mat[,16:1],
        col=gray(seq(1,0,l=12)),...)
  invisible(
    if (!use.first){
      title(x[1])
    }else{
    }
  )  
  #col=gray(seq(1,0,l=2)))
}
```


## 2. Local MDS for ZERO digits

```{r PART B reading data}
zip.train <- read.table("zip.train")
zip.train.0 <- zip.train[zip.train[1] == 0, -1]
```

```{r PART B-2 a}
row.names(zip.train.0) <- NULL
dist.zip.0 <- dist(zip.train.0)

k <- 5
tau <- .05
q<-2 # 2-dim config

conf0 <- stats::cmdscale(dist.zip.0, k=q)

if (!file.exists("lmds1.RData")) {
  lmds.S.res <- lmds(as.matrix(dist.zip.0), init=conf0, ndim=q, k=k, tau=tau, itmax = 1000)
} else {
  load("lmds1.RData")
}

conf.lmds.S.res <- lmds.S.res$conf
```

```{r PART B-2 a store, include=FALSE}
if (!file.exists("lmds1.RData")) {
  save(lmds.S.res, file="lmds1.RData")
}
```

```{r PART B-2 a2, out.width="90%", fig.align = "center"}
plot(conf.lmds.S.res, main=paste0("Local MDS, k=",k,", tau=",tau))
```
We've selected 9 points to encompass the data's variability. These points are marked in red, and each one has assigned a number in order to recognized each point with its representation using the *plot.zip* function.

```{r PART B-2 b, out.width="90%", fig.align = "center"}
#filt <- as.data.frame(conf.lmds.S.res)
#filt2 <- subset(filt, V1 > 40 & V1 < 55 & V2 < -7 & V2 > -14)
selected.points <- c(443, 575, 474, 291, 905, 731, 644, 242, 201)
aux <- conf.lmds.S.res[-selected.points,]
aux2 <- conf.lmds.S.res[selected.points,]

plot(aux, main=paste0("Local MDS, k=",k,", tau=",tau), col=8)
points(aux2, pch = 16, col=2)
text(aux2[,1],aux2[,2],1:9, pos=3, col=2, font = 2, cex = 1.5)
```

We can observe 2 patterns in the printed zeros below. The first one (the y-axis) is related to thickness of the stroke, positive values are thicker (the higher the value, the higher the stroke) and vice versa (lower values result in thinner strokes). On the other hand, the x-axis could be associated with the roundness of the number (negative values resemble an oval shape, while positive values resemble a circle shape). Furthermore, it appears that positive values in x-axis tend to result in unfinished strokes, like fast writing.


```{r PART B-2 b zero plots, fig.show="hold", out.width="32%"}
for (i in 1:length(selected.points)) {
  index <- selected.points[i]
  plot.zip(zip.train.0[index,], use.first=TRUE)
  title(i, cex.main = 2.5)
}

```


```{r PART B-2 c LCMC function, include=FALSE}
LCMC <- function(D1,D2,Kp){
  D1 <- as.matrix(D1)
  D2 <- as.matrix(D2)
  n <- dim(D1)[1]
  N.Kp.i <- numeric(n)
  for (i in 1:n){
    N1.i <- sort.int(D1[i,],index.return = TRUE)$ix[1:Kp]
    N2.i <- sort.int(D2[i,],index.return = TRUE)$ix[1:Kp]
    N.Kp.i[i] <- length(intersect(N1.i, N2.i))
  }
  N.Kp<-mean(N.Kp.i)
  M.Kp.adj <- N.Kp/Kp - Kp/(n-1)
  
  return(list(N.Kp.i=N.Kp.i, M.Kp.adj=M.Kp.adj))
}
```



```{r PART B-2 c}
#dist.zip.0 <- dist(zip.train.0)
q <- 2
Kp <- 10

K <- c(5,10,15)
Tau <- c(.1,.5,1)

if (!file.exists("lmds_k_tau.RData")) {
  LC <- matrix(0,nrow=length(K),ncol=length(Tau))
  lmds.k.tau <- array(vector("list",1),dim=dim(LC))
  
  for (i in 1:length(K)){
    for (j in 1:length(Tau)){
      lmds.k.tau[[i,j]] <- lmds(as.matrix(dist.zip.0), init=conf0, 
                                ndim=q, k=K[i], tau=Tau[j], itmax=1000)$conf
      D2.k.tau <- dist(lmds.k.tau[[i,j]])
      LC[i,j] <- LCMC(dist.zip.0,D2.k.tau,Kp)$M.Kp.adj
    }
    
  }
} else {
  load("lmds_k_tau.RData")
  load("LC.RData")
}


ij.max <- arrayInd(which.max(LC),.dim=dim(LC))
k.max <- K[ij.max[1]] 
tau.max <- Tau[ij.max[2]] 
lmds.max <- lmds.k.tau[[ij.max[1],ij.max[2]]]

print(paste0("k.max=",k.max,"; tau.max=",tau.max))
```
```{r PART B-2 c store, include=FALSE}
if (!file.exists("LC.RData")) {
  save(LC, file="LC.RData")
}

if (!file.exists("lmds_k_tau.RData")) {
  save(lmds.k.tau, file="lmds_k_tau.RData")
}
```



```{r PART B-2 c plot, out.width="90%", fig.align = "center"}
selected.points <- c(344, 962, 221, 530, 399, 440)
aux <- lmds.max[-selected.points,]
aux2 <- lmds.max[selected.points,]

plot(aux, main=paste0("Local MDS, k=",k.max,", tau=",tau.max), col=8)
points(aux2, pch = 16, col=2)
text(aux2[,1],aux2[,2],1:6, pos=2, col=2, font = 2, cex = 1.5)
```

```{r PART B-2 c zero plot, fig.show="hold", out.width="30%"}
for (i in 1:length(selected.points)) {
  index <- selected.points[i]
  plot.zip(zip.train.0[index,], use.first=TRUE)
  title(i, cex.main=2.5)
}
```

## 3.
```{r PART B-3 a}


```



## 4.
```{r PART B-4 a}
#row.names(zip.train.0) <- NULL
zip_matrix <- as.matrix(zip.train.0)
tsne_out <- Rtsne(zip_matrix, dims=2, pca=FALSE,perplexity=40,theta=0.0)
```

```{r PART B-4 a plot, out.width="90%", fig.align = "center"}
plot(tsne_out$Y)
```
Idem as before
```{r PART B-4 a plot, out.width="90%", fig.align = "center"}
#filt <- as.data.frame(tsne_out$Y)
#filt2 <- subset(filt, V1 > 35 & V1 < 40 & V2 > -5 & V2 < 0)
selected.points <- c(443, 575, 474, 139, 489, 732, 160, 242, 699 )
tsne_out.no.selected <- tsne_out$Y[-selected.points,]
tsne_out.selected <- tsne_out$Y[selected.points,]

plot(tsne_out.no.selected, col=8)
points(tsne_out.selected, pch = 16, col=2)
text(tsne_out.selected[,1],tsne_out.selected[,2],1:9, pos=2, col=2, font = 2, cex = 1.5)
```
```{r PART B-4 a zero plot, fig.show="hold", out.width="30%"}
for (i in 1:length(selected.points)) {
  index <- selected.points[i]
  plot.zip(zip.train.0[index,], use.first=TRUE)
  title(i, cex.main=2.5)
}
```

```{r PART B-2 c}
#dist.zip.0 <- dist(zip.train.0)
q <- 2
Kp <- 10

K <- c(5,10,15)
Tau <- c(.1,.5,1)

if (!file.exists("lmds_k_tau.RData")) {
  LC <- matrix(0,nrow=length(K),ncol=length(Tau))
  lmds.k.tau <- array(vector("list",1),dim=dim(LC))
  
  for (i in 1:length(K)){
    for (j in 1:length(Tau)){
      lmds.k.tau[[i,j]] <- lmds(as.matrix(dist.zip.0), init=conf0, 
                                ndim=q, k=K[i], tau=Tau[j], itmax=1000)$conf
      D2.k.tau <- dist(lmds.k.tau[[i,j]])
      LC[i,j] <- LCMC(dist.zip.0,D2.k.tau,Kp)$M.Kp.adj
    }
    
  }
} else {
  load("lmds_k_tau.RData")
  load("LC.RData")
}


ij.max <- arrayInd(which.max(LC),.dim=dim(LC))
k.max <- K[ij.max[1]] 
tau.max <- Tau[ij.max[2]] 
lmds.max <- lmds.k.tau[[ij.max[1],ij.max[2]]]

print(paste0("k.max=",k.max,"; tau.max=",tau.max))
```

## 5.
```{r PART B-5}


```




