---
title: "Smoothing and regression splines"
author: "Biel Caballero Vergés, Svenja Menzenbach and Kleber Enrique Reyes Illescas"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r loading libraries, include=FALSE}
library(sm)
```

### Load data
```{r loading data, include=FALSE}
load("bikes.Washington.Rdata")
attach(bikes)
```

## 1. Consider the nonparametric regression of cnt as a function of instant. Estimate the regression function m(instant) of cnt as a function of instant using a cubic regression spline estimated with the R function smooth.splines and choosing the smoothing parameter by Generalized Cross Validation.
```{r}
sm.sp <- smooth.spline(x = instant, y = cnt, 
                         cv = FALSE, all.knots = FALSE)
sm.sp

# Number of knots
sm.sp$fit$nk-2
```

### a) Which is the value of the chosen penalty parameter $\lambda$?
The value of $\lambda$ is 1.005038e-07.

### b) Which is the corresponding equivalent number of degrees of freedom df?
The corresponding equivalent number of degrees of freedom (df) is 93.34091.

### c) How many knots have been used?
We have used 134 knots.

### d) Give a graphic with the scatter plot and the estimated regression function $\hat{m}$(instant).
```{r}
plot(instant, cnt, col=8)
#abline(v=sm.sp$fit$min+sm.sp$fit$knot*sm.sp$fit$range, col=8, lty=2)
lines(sm.sp, col=2, lwd=2)
```


## 2. The script IRWLS logistic regression.R includes the definition of the function logistic.IRWLS.splines performing nonparametric logistic regression using splines with a IRWLS procedure. The basic syntax is the following: logistic.IRWLS.splines(x=..., y=..., x.new=..., df=..., plts=TRUE) where the arguments are the explanatory variable x, the 0-1 response variable y, the vector x.new of new values of variable x where we want to predict the probability of y being 1 given that x is equal to x.new, the equivalent number of parameters (or model degrees of freedom) df, and the logical plts indicating if plots are desired or not. Define a new variable cnt.5000 taking the value 1 for days such that the number of total rental bikes is larger than or equal to 5000, on 0 otherwise.
```{r}
source("IRWLS_logistic_regression.R")
cnt.5000 <- as.numeric(cnt >= 5000)
```

### a) Use the function logistic.IRWLS.splines to fit the non-parametric binary regression cnt.5000 as a function of the temperature, using df=6. In which range of temperatures is Pr(cnt >= 5000|temp) larger than 0,5?
```{r}
# Sort data according to x
x <- temp
y <- cnt.5000
sx <- sort(x,index.return =TRUE)
x <- sx$x
y <- y[sx$ix]

IRWLS.sp <- logistic.IRWLS.splines(x=x, y=y, x.new = x, df=6)

plot(x, y)
lines(x,IRWLS.sp$fitted.values,col=2)
abline(h=0.5, col=8, lty=2)

x.05 <- x[as.numeric(IRWLS.sp$predicted.values >= 0.5) == 1]
x.min <- min(x.05)
x.max <- max(x.05)
print(sprintf("min: %f,  max: %f", x.min, x.max))
```
Looking at the returned prediction, the temperatures between 21.7° and 32.4° have Pr(cnt >= 5000|temp) larger than 0.5. We can also check it looking at the plot.

### b) Choose the parameter df by k-fold log-likelihood cross validation with k = 5 and using df.v = 3:15 as the set of possible values for df.
```{r}
k.fold.ll.cv <- function(x,y,df,k=5){
    n <- length(x)
    Ik <- floor((0:(n-1))/(n/k))+1
    cum_sum <- 0
  
    for (i in (1:k)){
        y.i <- y[Ik==i]
        pred <- logistic.IRWLS.splines(x[Ik!=i], y[Ik!=i], 
                                       x.new=x[Ik==i], df=df)$predicted.values
        #cum_sum <- cum_sum + sum((y.i-pred)^2)
        cum_sum <- cum_sum + sum(y.i*log(pred/(1-pred)) + log(1-pred))
    }
    k.cv <- cum_sum/n
    return(k.cv)
}

```

```{r}
df.k.fold.ll.cv <- function(x,y,df.v,k=5){
    n <- length(x)
    perm <- sample(1:n)
    xperm <- x[perm]
    yperm <- y[perm]
  
    k.cv <- df.v*0
    for (i in (1:length(df.v))){
        df <- df.v[i]
        k.cv[i] <- k.fold.ll.cv(x=xperm, y=yperm, df, k)
    }
    return(list(k=k,df.v=df.v,k.cv=k.cv))
}
```



```{r}
df.v <- 3:15
result <- df.k.fold.ll.cv(x,y,df.v)

selected_df <- result$df.v[which.max(result$k.cv)]

plot(result$df.v, result$k.cv)
lines(result$df.v, result$k.cv)
abline(v = selected_df, col="8", lty=2)

```

```{r}
IRWLS.sp <- logistic.IRWLS.splines(x=x, y=y, x.new = x, df=selected_df)

plot(x, y)
lines(x,IRWLS.sp$fitted.values,col=2)
abline(h=0.5, col=8, lty=2)
```
