---
title: "Interpretability and Explainability in Machine Learning"
author: "Biel Caballero Vergés, Svenja Menzenbach and Kleber Enrique Reyes Illescas"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r loading libraries, include=FALSE}
library(readxl)
library(ranger)
library(vip)
library(gridExtra)
library(DALEX)
library(DALEXtra)
library(lime)
library(iml)
library(localModel)
# library(fastshap) # Attention! It re-define "explain" from DALEX
if (require(ghostvar)){library(ghostvar)}
library(mgcv)
library(grid)
library(ggplot2)
```

# Data preperation
```{r}
concrete <- as.data.frame(read_excel("Concrete_Data.xls"))
DescVars <- names(concrete)
names(concrete) <- c("Cement","Slag","FlyAsh","Water","Superplast", "CoarseAggr","FineAggr","Age","Strength")
```

```{r}
set.seed(42)

concrete_copy <- concrete

sample <- sample(nrow(concrete), 700)
train_set <- concrete_copy[sample,]
test_set <- concrete_copy[-sample,]

head(train_set)
head(test_set)
```

# 1. Fit a Random Forest
a. Compute the Variable Importance by the reduction of the impurity at the splits defined by each variable.
```{r}
model_rf_imp <- ranger(
  Strength ~ .,
  data = train_set, 
  importance='impurity'
)
print(model_rf_imp)

```

b. Compute the Variable Importance by out-of-bag random permutations.
```{r}
model_rf_perm <- ranger(
  Strength ~ .,
  data = train_set, 
  importance='permutation'
)
print(model_rf_perm)
```

c. Do a graphical representation of both Variable Importance measures.
```{r}
rf_imp_vip <- vip(model_rf_imp)
rf_perm_vip <- vip(model_rf_perm)
grid.arrange(rf_imp_vip, rf_perm_vip, ncol=2, top="Left: Reduction in impurity at splits. Right: Out-of-bag permutations")
```

d. Compute the Variable Importance of each variable by Shapley Values.
```{r}
y_hat<-function(model, data){
  predict(model, newdata=data)
}

rf_shapley <- vip(model_rf_imp, method = "shap",
                  pred_wrapper = y_hat,
                  train = train_set,  # argument 'train' must be provided
                  newdata = test_set[,-9],
                  num_features = 8)
```

# 2. Fit a linear model and a gam model.
a. Summarize, numerically and graphically, the fitted models.
```{r}
lm_strength <- lm(Strength ~ ., data = train_set)
(summ_lm_strength <- summary(lm_strength))
```
```{r}
gam_strength <- gam(Strength ~ s(Cement) + s(Slag) + s(FlyAsh) + s(Water) + s(Superplast) + s(CoarseAggr) + s(FineAggr) + s(Age), # change model?
                 data = train_set)
(summ_gam_strength <- summary(gam_strength))
```
```{r}
plot(lm_strength)
plot(gam_strength)
```

b. Compute the Variable Importance by Shappley values in the linear and gam fitted models. Compare your results with what you have learned before.

```{r}
lm_strength_shapley <- vip(lm_strength, method="shap",
                  pred_wrapper=predict.lm,
                  train=train_set, # train set must be specified
                  newdata=test_set[,-9], 
                  num_features = 8,
                  exact=TRUE)

plot(lm_strength_shapley)
```
```{r}
gam_strength_shapley <- vip(gam_strength, method="shap",
                  pred_wrapper=predict.gam,
                  train=train_set, # train set must be specified
                  newdata=test_set[,-9],
                  num_features = 8,
                  exact=TRUE)

plot(gam_strength_shapley)
```

# 3. Relevance by Ghost Variables
Compute the relevance by ghots variables in the three fitted models.
```{r}
source("relev.ghost.var.R")
Rel_Gh_Var <- relev.ghost.var(model=gam_strength, 
                              newdata = test_set[, -9],
                              y.ts = test_set[, 9],
                              func.model.ghost.var = lm
)
plot.relev.ghost.var(Rel_Gh_Var,n1=500,ncols.plot = 3)

```

# 4. Global Importance Measures and Plots using the library DALEX
a. Compute Variable Importance by Random Permutations
```{r}
explainer_rf <- explain.default(model = model_rf_imp,  
                               data = test_set[, -9],
                               y = test_set[, 9], 
                               label = "Random Forest")
```

b. Do the Partial Dependence Plot for each explanatory variable.
```{r}
PDP_rf <- model_profile(
  explainer=explainer_rf,
  variables = NULL,  # All variables are used
  N = NULL, # All available data are used
  groups = NULL,
  k = NULL,
  center = TRUE,
  type = "partial" #  partial, conditional or accumulated
)

plot(PDP_rf, facet_ncol=2)
```
For Cement and CoarseAggr, the predicted Strength initially rises with an increase in these materials but eventually decreases after reaching an optimal point. This suggests an optimal quantity for both components, as excessive use could diminish Strength. On the other hand, FlyAsh and Slag show a consistent increase in predicted Strength with higher quantities, implying their positive impact on concrete Strength.     

c. Do the Local (or Conditional) Dependence Plot for each explanatory variable.
```{r}
CDP_rf <- model_profile(
  explainer=explainer_rf,
  variables = NULL,  # All variables are used
  N = NULL, # All available data are used
  groups = NULL,
  k = NULL,
  center = TRUE,
  type = "conditional" #  partial, conditional or accumulated
)

plot(CDP_rf, facet_ncol=2)
```
While cement and coarse aggregate initially boost concrete strength, their impact diminishes at higher proportions. Optimal dosages exist for these materials, as exceeding them may impair strength. In contrast, fly ash and slag consistently enhance concrete strength with increasing amounts. 


# 5. Local explainers with library DALEX
Choose two instances in the the test set, the prediction for which we want to explain:    
• The data with the lowest value in Strength.     
• The data with the largest value in Strength.    
For these two instances, do the following tasks for the fitted random forest.    

```{r}
lowestStrength = concrete[which.min(concrete$Strength), ]
highestStrength = concrete[which.max(concrete$Strength), ]
```

a. Explain the predictions using SHAP.
```{r}
bd_rf <- predict_parts(explainer = explainer_rf,
                 new_observation = lowestStrength,
                            type = "shap")

bd_rf
plot(bd_rf)
```
This plot shows that the features FineAggr, Cement, Superplast and Slag have the biggest impact (positively) and CoarseAggr negatively.


```{r}
bd_rf <- predict_parts(explainer = explainer_rf,
                 new_observation = highestStrength,
                            type = "shap")

bd_rf
plot(bd_rf)
```
This plot shows that all features have a good contribution towards Strength.

b. Explain the predictions using Break-down plots.
```{r}
bd_rf <- predict_parts(explainer = explainer_rf,
                 new_observation = lowestStrength,
                            type = "break_down")

bd_rf
plot(bd_rf)
```
Here the plot shows that Cement and Superplast have a significant impact on the Strength. This means that we can focus on optimizing these two input variables to achieve the desired Strength.


```{r}
bd_rf <- predict_parts(explainer = explainer_rf,
                 new_observation = highestStrength,
                            type = "break_down")

bd_rf
plot(bd_rf)
```
This plot shows that the predicted concrete strength increases with increasing CoarseAggr proportions up to a point of around 50%.    


c. Explain the predictions using LIME.
```{r}
bd_rf <- predict_surrogate(explainer = explainer_rf,
                 new_observation = lowestStrength,
                            type = "localModel")

bd_rf
plot(bd_rf)
```
This plot that Cement, Water, Superplast and FineAggr have the biggest positive impact while Slag the biggest negative impact.


```{r}
bd_rf <- predict_surrogate(explainer = explainer_rf,
                 new_observation = highestStrength,
                            type = "localModel")

bd_rf
plot(bd_rf)
```
The plot shows all having a big impact, but Cement and Water being the top ones.


d. Do the Individual conditional expectation (ICE) plot, or ceteris paribus plot
```{r}
cp_rf <- predict_profile(explainer = explainer_rf,
                 new_observation = lowestStrength)

cp_rf
plot(cp_rf,facet_ncol=2)
```
The plots show that the predicted concrete strength increases with increasing cement content, fine aggregate content, superplasticizer content, and slag content. However, the predicted concrete strength initially increases with increasing coarse aggregate content, but eventually reaches a peak and then declines.


```{r}
cp_rf <- predict_profile(explainer = explainer_rf,
                 new_observation = highestStrength)

cp_rf
plot(cp_rf,facet_ncol=2)
```
The findings indicate that the predicted concrete strength increases linearly with increasing cement content and fine aggregate content. In contrast, the predicted concrete strength increases non-linearly with increasing superplasticizer content and slag content.    


e. Plot in one graphic the Individual conditional expectation (ICE) plot for variable Age for eachcase in the test sample. Add the global Partial Depedence Plot.
```{r}
mp_rf <- model_profile(explainer = explainer_rf,
  variables = "Age",
  N = 100,
  type = "partial"
)

plot(mp_rf, geom = "profiles") +  
  ggtitle("Ceteris-paribus and partial-dependence profiles for Age") 
```
The plot shows that the predicted Strength of concrete generally increases with increasing Age, but the relationship is complex and non-linear. The average effect of Age on Strength is positive, but the effect diminishes at higher ages.
